---
title: "Hypothesis Testing"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
    code-fold: show
    self-contained: true
tbl-cap-location: bottom
execute:
  echo: true
---

```{r}
#| include: false
library(ggplot2)

theme_set(
    theme_minimal(base_size = 16)
)
update_geom_defaults("line", list(size = 1.1))
```

# Motivation

## Motivating Hypothesis Testing Framekwork

Suppose your buddy claims he shoots 90% from the free throw line.  Since we are
all statisticians, we a) don't believe them, and b) insist upon testing their claim
empirically.  So we collect some data.  They step up and start shooting.  At
what point do we reject their claim?

## Motivating Hypothesis Testing Framekwork

Implicitly, we used logical framework to evaluate our buddy's claim.  Let's
unpack that framework and give it a name.


* Establish two hypotheses

    * they shoot as well as they say they do, $p = 0.9$
    * they do not shoot as well as they claim, $p < 0.9$.

* Collect data

    * He took $n$ shots from the line.

* Analyze the data

    * Estimated $p$ with $\hat{p}$
    * Determined the probability of observing $\hat{p}$ if indeed he shoots as well as he claims.

* Make a conclusion based on your analysis

    * Given his claim, if $\hat{p}$ seems too unlikely, the claim is probably not true.


## Hypothesis Test, another example

Suppose Giggle wants to test their browser Chime.  They sample 1014 randomly
selected websites and make a simple decision, this website was displayed
properly on Chime or it was not.  They found that 984 websites displayed
correctly.  Test whether or not Chime displays 99% of webpages correctly, and
compare your conclusion to a confidence interval.  Choose $\alpha = 0.05$.

* Establish two hypotheses
* Collect data
* Analyze data $\Rightarrow$ calculate z-score-like thing
* Make decision


## Hypothesis Test, another example continued

Hypothesis Test:

$$H_0: p = .99 \quad \text{ versus } \quad H_1: p \ne .99$$

```{r}
n <- 1014
x <- rep(0, n)
x[1:984] <- 1
t.test(x, mu = 0.99, conf.level = 0.99) # or
# prop.test(sum(x), n) # if you're being picky
# compare p-value to α = 0.05
```

# Framework

## Hypothesis Testing Framework

We call this framework **hypothesis testing**.  Let's rephrase hypothesis
testing into the language of statistics.

* State **hypotheses**, null $H_0$ and alternative $H_1$.
* Collect data.
* Calculate **test statistic** and **p-value**.
* Conclude by comparing p-value to **level of significance**.


## hypotheses

The null and alternative hypotheses generally follow some conventions.

* $H_0$ and $H_1$ are statements about *population parameters*

* $H_0$ declares the parameter of interest to be *equal* to some value.

    * $H_0: p = 0.99$ or $H_0: \mu = 10$

* $H_1$ declares the (same) parameter of interest to be less than, greater than,
    or not equal to the same value in the null hypothesis $H_0$ -- the
    researcher chooses one before collecting data, let alone conducting the
    test.

    * $H_1: p \ne 0.99$ or $H_1: \mu \; \{<,>,\ne\} \; 10$


## test statistic

::: {.nonincremental}
* A **test statistic** is a summary statistic that is particularly useful for
  evaluating a hypothesis test or calculating the **p-value**.
:::

```{r}
t.test(x, mu = 0.99, conf.level = 0.99)
```


## p-value

::: {.nonincremental}
* The **p-value** is a probability, which provides weights evidence against
$H_0$ (but never provides evidence in favor of anything, let alone $H_1$)
:::

:::{.callout-note}

It's too easy to think you proved something when p-value $< \alpha$, however
statistics rarely *proves* anything.  At best, statistics, via p-values,
provides evidence against a specific conclusion, namely $H_0$.

:::


## p-value

::: {.nonincremental}
* **p-value**. The probability of observing the test statistic we did, or
  something more extreme[^1], assuming the null hypothesis is true.
:::

::: {.callout-note}
Because the p-value is a probability, there's two sides to it.  The other side
results in an error in decision making.
:::

[^1]: smaller, larger, or both, depending on the alternative hypothesis.

## p-value, in picture

For an alternative hypothesis of $\ne$, namely $H_1: mu \ne 0$,

```{r}
#| echo: false

p <- ggplot(data.frame(x=c(-4,4)), aes(x)) + stat_function(fun=dnorm) +
    labs(x="", y="")
d <- ggplot_build(p)$data[[1]]
p + geom_area(data=subset(d, x>qnorm(0.975)),
              aes(x=x, y=y), fill="black", colour="black") +
    geom_area(data=subset(d, x<qnorm(0.025)),
              aes(x=x, y=y), fill="black", colour="black") +
    theme(axis.text.y=element_blank(),
          axis.ticks=element_blank())
```


## level of significance

* We define a largest level at which we are willing to incorrectly conclude.  We
  call this value the level of significance, and give it the symbol
  $\alpha$[^1].

* **level of significance**. The largest probability of incorrectly rejecting
  $H_0$ when in fact $H_0$ is true.

[^1]: Common values of $\alpha$ are $0.05$ and $0.01$.


## Hypothesis Testing, conclusions via p-value

We evaluate hypotheses by comparing the p-value to the significance level,
$\alpha$.  If our sample statistic (observed data) is so unusual with respect to
the null hypothesis that it casts doubt on the validity of $H_0$, then we have
*some evidence against* $H_0$ (but not confirming $H_1$).


## evidence against H_0

Hypothesis testing never confirms anything, it provides some evidence against
the null hypothesis.

When

::: {.nonincremental}
* p-value $\leq \alpha \Rightarrow$ reject $H_0$
* p-value $> \alpha \Rightarrow$ fail to reject $H_0$
:::

::: {.callout-note}
Despite the overly cautious words above (some, evidence, probably), the world of
statistics continues to use the strong phrases "reject" and "fail to reject".
:::


## Last Example, finch beak height

Consider again Darwin's finch data set.  Formally test that the mean beak height
is equal to 9mm versus not equal to, at $\alpha = 0.05$.

```{r}
url <- "https://raw.githubusercontent.com/roualdes/data/master/finches.csv"
finch <- read.csv(url)
```

## Last Example, finch beak height

* $H_0: \mu = 9$ versus $H_1: \mu \ne 9$;  $\alpha = 0.05$.
* calculate test statistic
* calculate p-value from test statistic
* conclude

```{r}
t.test(finch$beakheight, mu = 9, conf.level = 0.95)
```

## Last Example, finch beak height

Given $H_0: \mu = 9$ and $H_1: \mu \ne 9$, we reject the null hypothesis because
the p-value $2.2e-16 < \alpha = 0.05$.  There is sufficient evidence to say that
the true population mean beak height of finches from the Galapagos islands is
not equal to 9mm.


# Decision Making is hard

## Hypothesis Testing, the decision

We conclude a formal hypothesis test by making a decision between $H_0$ and
$H_1$.  But did we decide correctly?


## Hypothesis Testing, decision errors

We essentially made a choice, but our choice could be correct or incorrect.

| |fail to reject $H_0$ | reject $H_0$ |
|-|---------------------|--------------|
|$H_0$ true| correct | **type 1 error**|
| $H_1$ true | **type 2 error** | correct|


## Hypothesis Testing, errors

* Type 1 Error
    * A **type 1 error** is rejecting the null hypothesis when $H_0$ is actually true.

* Type 2 Error
    * A **type 2 error** is failing to reject the null hypothesis when the
      alternative is actually true.


## Hypothesis Testing, error trade offs

OS4 Example 5.26.  How could we reduce the Type 2 Error rate in US courts?  What
influence would this have on the Type 1 Error rate?

| | found innocent | found guilty |
|-|---------------------|--------------|
|didn't commit crime| correct | **type 1 error**|
|did commit crime  | **type 2 error** | correct|

* Lower Type 2 Error by convicting more people; change standards from "beyond a
  reasonable doubt" to "beyond a little doubt".
* What then of wrongful convictions, i.e.\ Type 1 Error rates?


## Hypothesis Testing, Connection to CIs

Confidence intervals can be used to conclude hypothesis tests if

::: {.nonincremental}
* two sided alternative
* $\alpha$ is equal to $1 - confidence$.
:::

Re finch beak height. So what if we tested $H_0: \mu = 13$ versus $H_1: \mu \ne
13$?  Would we reject or fail to reject $H_0$?


## Take Away

Hypothesis testing


* formalizes someone's intuition and logical thought process
* relies on new words: **hypotheses**, **test statistic**, **p-value**, **level of significance**
* right or wrong, is ubiquitous in statistcs
* connects to confidence intervals in specific ways
