---
title: "Linear Regression Inference"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
    code-fold: true
    self-contained: true
tbl-cap-location: bottom
execute:
  echo: true
---

```{r}
#| include: false
library(ggplot2)

theme_set(
    theme_minimal(base_size = 16)
)
update_geom_defaults("line", list(size = 1.1))
```

# Recap

## recap: central limit theorem

The Central Limit Theorem says, "If our sample size is large enough, the sample
mean will be approximately Normally distributed."


## recap: confidence intervals

From the CLT, we can approximate confidence intervals from an approximate
*sampling distribution*.

```{r}
#| echo: false

p <- ggplot(data.frame(x=c(-4,4)), aes(x)) + stat_function(fun=dnorm) +
    labs(x="", y="")
d <- ggplot_build(p)$data[[1]]
p + geom_area(data=subset(d, x<qnorm(0.975) & x>qnorm(0.025)),
              aes(x=x, y=y), fill="black", colour="black") +
    theme(axis.text.y=element_blank(),
          axis.ticks=element_blank())
```

## recap: hypothesis tests

From the CLT, we can approximate area in tails (p-values) from an approximate
*sampling distribution*.


```{r}
#| echo: false

p <- ggplot(data.frame(x=c(-4,4)), aes(x)) + stat_function(fun=dnorm) +
    labs(x="", y="")
d <- ggplot_build(p)$data[[1]]
p + geom_area(data=subset(d, x>qnorm(0.975)),
              aes(x=x, y=y), fill="black", colour="black") +
    geom_area(data=subset(d, x<qnorm(0.025)),
              aes(x=x, y=y), fill="black", colour="black") +
    theme(axis.text.y=element_blank(),
          axis.ticks=element_blank())
```

## recap: linear regression


Linear regression is a method to fit a line through a scatter plot of data in a
"best" sense.  Often, interest lies in the relationship between the
explanatory and the response variable.


## recap: linear regression, plot

```{r}
url <- "https://raw.githubusercontent.com/roualdes/data/master/elmhurst.csv"
elmhurst <- read.csv(url)
ggplot(data = elmhurst, aes(family_income, gift_aid),
       ylab = "Gift aid ($1000)",
       xlab = "Family income ($1000)") +
    geom_point() +
    stat_smooth(method = "lm", se = FALSE)
```

## recap: linear regression, code

```{r}
elmReg <- lm(gift_aid ~ family_income, data=elmhurst)
## summary(elmReg) ## RStudio
## yhat <- fitted(elmReg)
## e <- residuals(elmReg)
beta <- coef(elmReg)
```

$$\widehat{aid} = `r round(beta[1], 2)` + `r round(beta[2], 2)` \times
family\_income$$

# Linear Regression Inference

## Estimating $\beta_0, \beta_1$

Linear regression estimates the population parameters $\beta_0$ (intercept) and
$\beta_1$ (slope), just like every other parameter we have estimated.  As
such, the estimators $\hat{\beta}_0$ and $\hat{\beta}_1$ of these parameters
have their own sampling distributions.

## Inference $\beta_0, \beta_1$

It turns out that the sampling distribution of $\hat{\beta}$ is approximately
normally distributed when the sample size is sufficiently large; CLT (again).

## Linear regression hypothesis tests

Hypothesis testing naturally follows.  The most common hypothesis test for
linear regression parameters is


\begin{align*}
    H_0: \quad & \beta = 0 \\
    H_1: \quad & \beta \ne 0
\end{align*}

with $\alpha = 0.05$.


## Standard Output

The hypothesis test above has a natural and informative interpretation in most
contexts.


```{r}
url <- "https://raw.githubusercontent.com/roualdes/data/master/elmhurst.csv"
elmhurst <- read.csv(url)
elmReg <- lm(gift_aid ~ family_income, data = elmhurst)
# summary(elmReg) # RStudio
```

## Standard Output

Be sure to understand and be able to find at least,

* p-values
* adjusted $R^2$

and knowing

* standard errors
* t values (test statistics)

will just make you sound smart.


## linear regression confidence intervals

If we can do hypothesis testing, we can do confidence intervals.  The function
`confint` in `R` is extremely helpful.

```{r}
#| eval: false

# use lm fitted model, elmReg from above
confint(elmReg) # default is 95%
confint(elmReg, level = 0.98) # can specify confidence
```

## linear regression predictions

We can make predictions with

```{r}
#| eval: false

# use lm fitted model, elmReg from above
predict(elmReg, newdata = data.frame(family_income = 50))
predict(elmReg,
        newdata = data.frame(family_income = 50),
        interval = "confidence",
        level = 0.98)
```


# $R^2$

## $R^2$

It is common to use the square of the (Pearson) correlation to explain the
strength of a linear fit.

The $R^2$ of a linear model describes the amount of variation in the response
variable $y$ that is explained by the least squares line on the explanatory
variable $x$.


## $R^2$, example

Using the data frame `elmhurst`, the correlation between gift aid and family
income is $R =$ `r round(with(elmhurst, cor(gift_aid, family_income)), 4)`.
Thus, $R^2 =$ `r round(with(elmhurst, cor(gift_aid, family_income)^2), 4)`.

We say `r round(with(elmhurst, 100*cor(gift_aid, family_income)^2), 2)`% of the
variation in `gift_aid` is explained by the least squares line on
`family_income`.


# Extrapolation

## Extrapolation, example

At age $8$, Shaquille O'Neal was 4'8".  At age 16, he was 6'8".  Can we use
these data to predict how tall Shaq is now that he is 43?

## Extrapolation, example

At age $8$, Shaquille O'Neal was 4'8".  At age 16, he was 6'8".  Can we use
these data to predict how tall Shaq is now that he is 50?

In eight years, Shaq grew 2 feet.  34 years later, Shaq should be an
additional 8'6" taller than he was at 16, thus 15'2".  Sound
reasonable?

## Extrapolation, example

At age $8$, Shaquille O'Neal was 4'8".  At age 16, he was 6'8".  Can we use
these data to predict how tall Shaq is now that he is 50?

In eight years, Shaq grew 2 feet.  34 years later, Shaq should be an
additional 8'6" taller than he was at 16, thus 15'2".  Sound
reasonable?

Note: Shaq is 7'1".


## Extrapolation, definition

Applying a model to values outside of the range of the original data is called
**extrapolation**.

:::{.callout-note}
Extrapolation is in general dangerous.  Sometimes it works, but not often, so
watch out.
:::


## Extrapolation, example

How much gift aid would a student expect to receive if their family
income\footnote{Don't forget family income is in units of $\$1000$.} was $\$1$
million?  Using our least squares line,

$$\widehat{aid} = `r round(beta[1], 2)` + `r round(beta[2], 2)` \times family\_income$$


* we'd estimate `r round(beta[1] + 1000*beta[2], 2)` thousand dollars.


## Take Away

* confidence intervals and hypothesis testing for linear regression
* predictions from linear regression
* $R^2$ to summarize the fit of a linear model
* extrapolation, predictions are more difficult outside the range of your data
