---
title: "ANOVA"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
    code-fold: true
    self-contained: true
tbl-cap-location: bottom
execute:
  echo: true
---

```{r}
#| include: false
library(ggplot2)

theme_set(
    theme_minimal(base_size = 16)
)
update_geom_defaults("line", list(size = 1.1))
```

# Recap

## recap: point estimates

Point estimates are random variables.  Random variables follow shapes, called
distributions.  Therefore, point estimates follow distributions (and have
shapes).

## recap: central limit theorem

The Central Limit Theorem says, "If our sample size is large enough, the sample
mean will be approximately Normally distributed."


## recap: confidence intervals

From the CLT, we can approximate confidence intervals

```{r}
#| echo: false

p <- ggplot(data.frame(x=c(-4,4)), aes(x)) + stat_function(fun=dnorm) +
    labs(x="", y="")
d <- ggplot_build(p)$data[[1]]
p + geom_area(data=subset(d, x<qnorm(0.975) & x>qnorm(0.025)),
              aes(x=x, y=y), fill="black", colour="black") +
    theme(axis.text.y=element_blank(),
          axis.ticks=element_blank())
```

## recap: hypothesis tests

From the CLT, we can approximate area in tails (p-values)


```{r}
#| echo: false

p <- ggplot(data.frame(x=c(-4,4)), aes(x)) + stat_function(fun=dnorm) +
    labs(x="", y="")
d <- ggplot_build(p)$data[[1]]
p + geom_area(data=subset(d, x>qnorm(0.975)),
              aes(x=x, y=y), fill="black", colour="black") +
    geom_area(data=subset(d, x<qnorm(0.025)),
              aes(x=x, y=y), fill="black", colour="black") +
    theme(axis.text.y=element_blank(),
          axis.ticks=element_blank())
```

## recap: two sample t-tests

The two sample $t$-test, compares the means of two groups with the hypothesis

$$
\begin{align*}
H_0: \quad & \mu_1 = \mu_2 \\
H_A: \quad & \mu_1 \ne \mu_2. \\
\end{align*}
$$

R's function `t.test()` did all the hard work for us.


# Analysis of Variance

## Analaysis of Variance

If there were three or more groups, the two sample $t$-test would not work.  We
could force the test on the data by comparing two groups at a time, but this has
dangerous implications.  We thus require a new statistical method, **analysis of
variance**.

* **Analysis of variance (ANOVA)** uses a single hypothesis to check whether the
  *means* across two or more groups are equal.

## ANOVA, hypothesis

The ANOVA hypothesis test for $k$ groups is

$$
\begin{align*}
H_0: \quad & \mu_1 = \mu_2 = \mu_3 = \ldots = \mu_k \\
H_A: \quad & \text{at least one mean is different.}
\end{align*}
$$

:::{.callout-note}
ANOVA tests equality of means across groups, despite its name.
:::


## ANOVA, hypothesis examples

With ANOVA you can compare the means by groups for many different data sets.


* mean batting average by position,
* mean movie budget by genre, (or year),
* mean CO2 intake by treatment, or location, (or concentration)
* mean rem sleep by conservation status,
* mean birth/body weight by family,
* ...


## ANOVA, picture

Let's visualize what is going on with ANOVA.  It starts
with box plots by groups -- draw more pictures on board.

```{r}
#| echo: false

df = data.frame(numvar = c(rnorm(35, -3), rnorm(35, 3), rnorm(35, 8)),
                fctr = rep(c("A", "B", "C"), each=35))

ggplot(data = df, aes(fctr, numvar)) +
    geom_boxplot()
```


## ANOVA, intuition

Analysis of variance tells us about means by groups, despite its name.  Large
variation amongst the groups relative to small variation within the groups
indicates different population means.

## ANOVA, intuition in picture

What do you think of these means -- think of variation within and amongst
groups?

```{r}
#| echo: false

df <- data.frame(y=c(rnorm(35, 0.1), rnorm(35, 0.05), rnorm(35, 0.15)),
                 f = rep(c("A", "B", "C"), each=35))
ggplot(data = df, aes(f, y)) +
    geom_boxplot() +
    stat_summary(fun.y=mean, colour="red", geom="point", size = 4)
```


## ANOVA, intution via picture

What do you think of these means -- think of variation within and amongst
groups?

```{r}
#| echo: false

df = data.frame(x=c(rnorm(35, 10), rnorm(35, -2), rnorm(35, 5)),
                f = rep(c("A", "B", "C"), each=35))
ggplot(data = df, aes(f, x)) +
    geom_boxplot() +
    stat_summary(fun.y=mean, colour="red", geom="point", size = 4)
```


## ANOVA, test statistic

ANOVA calculates one fraction based on two numbers, variation amongst (between)
groups and variation within groups.  These two numbers are generally referred to
as mean square values.


* mean squared amongst groups
    * **MSG** is a strictly positive measure of the variation across all groups.

* mean squared error/residuals
    * **MSE** is a strictly positive measure of the variation within groups.

* Test statistic
    * $F = \frac{MSG}{MSE}$


# First Example

## Baseball example, hypotheses

Are baseball players paid on average differently by position?

$$
\begin{align*}
  H_0: \quad & \mu_{catcher} = \mu_{dh} = \mu_{first} = \ldots = \mu_{third} \\
  H_A: \quad & \text{at least one mean salary is different.}
\end{align*}
$$

with $\alpha = 0.05$.

## Baseball example, plot

Load the data and make a plot.

```{r}
url <- "https://raw.githubusercontent.com/roualdes/data/master/mlb.csv"
mlb <- read.csv(url)
suppressMessages(library(tidyverse))
ggplot(data = mlb, aes(position, salary)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```


## Baseball example, R code

The R code to run ANOVA

```{r}
#| output: false

linear_model <- lm(salary ~ position, data = mlb)
anova(linear_model)
```


## Baseball example, R code details

The tilde $\sim$ is meant to be read as, predict the left hand side with the
right hand side.  Hence, we read the following as, predict salary by different
levels of position.


```{r}
#| eval: false
linear_model <- lm(salary ~ position, data = mlb)
```

## Baseball example, R output

The degrees of freedom, F-statistic, and p-value are the most important pieces
of information to extract from an ANOVA table.

```{r}
anova(linear_model)
```


## Baseball example, conclusion

Because the p-value $= `r round(anova(linear_model)[[5]][1], 4)` < \alpha =
0.05$ we reject the null in favor of the alternative.  There is sufficient
evidence to claim that the population of mean salaray of baseball players
probably varies by position.


# Multiple Comparisons

## Multiple Comparisons, the problem

Say you've got some data and you want to test equality of the means. What to do?
ANOVA!

```{r}
ggplot(data = mlb, aes(position, salary)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Multiple Comparisons, the problem

What not to do?  Immediately compare all pairwise combinations of the $k = 9$
groups.  That would result in `r choose(9, 2)` tests, and an increase in your
Type I Error rate.


##  Multiple Comparisons, how bad a problem

Suppose you chose a level of significance of $\alpha = 0.05$.  What's the
probability of observing at least one significant result just by chance?

This is just a binomial distribution, $X \sim$ Binomial$(28, 0.05)$, and we want to know $P(X \ge 1)$.

```{r}
#| echo: false
1 - pbinom(0, 28, 0.05)
```


## Multiple Comparisons, a solution

Many solutions exist.  A simple one, in R, is Tukey's honest significant
difference (HSD).  Only use this if you have first run ANOVA and rejected $H_0$.

```{r}
#| eval: false

model <- lm(salary ~ position, data = mlb) # if reject H0
TukeyHSD(aov(model)) # do in R
```


# Last Example

## Carnivora Brain Weights by Family, plot

Consider brain weights of the Families of the order Carnivora.  Let's compare
mean brain weights by Family.

```{r}
suppressMessages({library(tidyverse)
  library(ape)
  data(carnivora)})
ggplot(data = carnivora, aes(Family, SB)) +
    geom_boxplot() +
    labs(y="Brain weights (g)") +
    theme(axis.text.x=element_text(angle=45, hjust=1))

```

## Carnivora Brain Weights by Family, not all groups

Since some groups don't have enough data, let's remove them.

```{r}
carnivs <- filter(carnivora,
                  !(Family %in% c("Ailuridae",
                                  "Procyonidae",
                                  "Viverridae")))
```

## Carnivora Brain Weights by Family, hypotheses

$$
\begin{align*}
  H_0: \quad & \mu_c = \mu_f = \mu_h = \mu_m = \mu_u \\
  H_A: \quad & \text{at least one mean is different.}
\end{align*}
$$

with $\alpha = 0.01$.


## Carnivora Brain Weights by Family, R code

```{r}
mod <- lm(SB ~ Family, data = carnivs)
anova(mod)
```


## Carnivora Brain Weights by Family, conclusion

Because the p-value is tiny, we reject $H_0$ in favor of the alternative.
There is sufficient evidence to say that at least one mean is different from the
rest.



## Carnivora Brain Weights by Family, post-hoc analysis

```{r}
#| eval: false

TukeyHSD(aov(mod), conf.level = 0.99) # do in R
```


## Take Away

* ANOVA breaks up the response variable into groups, chosen by you.
* ANOVA essentially calculates group means
    * average error within groups and average error amongst the groups.
* Understand the general ANOVA table
* Conclusions from ANOVA tables are easy
    * compare p-value to level of significance
* Interpretation of such conclusions is trickier, practice
