---
title: "Quiz 02"
format: html
editor: visual
---

```{r}
library(ggplot2)
suppressMessages(library(dplyr))
df <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/penguins.csv")
```

Consider [this](https://raw.githubusercontent.com/roualdes/data/refs/heads/master/penguins.csv) data set about penguins. We'll use the variables `species`, `bill_length_mm`, and `body_mass_g`. The bill lengths are measured in millimeters and body mass is measured in grams.

```{r fig.width=5, fig.height=2, fig.align='center'}
ggplot(data = df, aes(species, bill_length_mm)) +
  geom_boxplot() +
  labs(x = "Species", y = "Bill length (mm)")
```

```{r}
fit <- lm(bill_length_mm ~ species, data = df)
anova(fit)
```

1.  Write appropriately matching hypotheses for the model (and plot) above.

2.  Use the output above to conclude the hypothesis test. Justify your conclusion, citing a p-value.

3.  What does your conclusion mean in English?

```{r}
summary(fit)
```

4.  Use the output above to calculate and interpret, in context of the data, an intercept for the Species Chinstrap.

```{r fig.width=5, fig.height=2, fig.align='center'}
ggplot(data = df, aes(bill_length_mm, body_mass_g)) + 
  geom_point() +
  labs(x = "Bill length (mm)", y = "Body mass (g)")
```

```{r}
fit <- lm(body_mass_g ~ bill_length_mm, data = df)
summary(fit)
```

5.  Write appropriately matching hypotheses for the intercept for the model (and plot) above.

6. Use the output above to conclude the hypothesis test. Justify your conclusion, citing a p-value.

7.  Interpret the intercept in context of the data.

8.  Interpret the slope in context of the data.

9.  What units does the slope have?

```{r}
confint(fit)
```

10.  Use the output above to interpret a confidence interval for a term of your choice in context of the data.

```{r}
predict(fit, newdata = data.frame(bill_length_mm = c(40, 10)))
```

11. Interpret a prediction in context of the data.

12. One of the two predictions is generally considered unreasonable. Which and why?
