---
title: "Central Limit Theorem"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
tbl-cap-location: bottom
execute:
  echo: true
---

```{r}
#| include: false
library(ggplot2)

theme_set(
    theme_minimal(base_size = 16)
)
update_geom_defaults("line", list(size = 1.1))
```

# Recap

## recap: point estimates

Point estimates are random variables.  Random variables follow shapes, called
distributions.  Therefore, point estimates follow distributions (and have
shapes).

## Central Limit Theorem

* The Central Limit Theorem says, "If our sample size is large enough, the
  sample mean will be approximately Normally distributed."

* The Central Limit Theorem says, if we have a collection of sample means, the
  shape (histogram) of this collection is basically Normal (unimodal and
  symmetric).


## CLT, via web app

Javascript web application to show [Central Limit
Theorem](http://onlinestatbook.com/stat_sim/sampling_dist/}{Central Limit
Theorem).


## CLT, definition (in Statistics)

The Central Limit Theorem says, "Under certain conditions, the sampling
distribution for the sample mean converges to the normal distribution as the
sample size increases."


## CLT, definition (in symbols)

When $n$ is sufficiently large,

$$\frac{\bar{X} - \mu}{\sigma_{\bar{X}}} \overset{\cdot}{\sim} N(0, 1).$$

As $n$ increases, the approximation improves.


## CLT's Assumptions

We should mention the conditions necessary for this to happen.

* independent observations,
* identically distributed, $X_i \sim \mathcal{F}$ for all $i$, and
* variance is finite, $\sigma^2 < \infty$.


## CLT, so what

The assumptions are really not that bad, so we can safely assume they hold in
many real world applications.  With that, so long as we use the mean, then we
can say, at least approximately, how the distribution of means is shaped -- even
if we never actually sample/calculate multiple means.


## Not Necessarily Normal Data

It doesn't matter if that data are from a Normal distribution or not; use mean =>
Central Limit Theorem.


## Example

We will use a sample of Darwin's finch data set [Swarth:1931].  Make a 98%
confidence interval about the beak height using Darwin's finches.

```{r}
# load data
url <- "https://raw.githubusercontent.com/roualdes/data/master/finches.csv"
finch <- read.csv(url)
# look at data in RStudio
```

## Example, continued

98% confidence interval for Darwin's finch data set.

```{r}
xbar <- mean(finch$beakheight)
std <- sd(finch$beakheight)
n <- length(finch$beakheight)
t <- qt(0.99, n - 1)
```

$$\bar{x} \pm t * \frac{std}{\sqrt{n}}$$

```{r}
xbar - t * std / sqrt(n) # lower
xbar + t * std / sqrt(n) # upper
```

## Example, simplified

```{r}
t.test(finch$beakheight, conf.level = 0.98)
```

## Revisit Web App

This time, pay attention to the width of the sampling distribution.

[Central Limit
Theorem web app](http://onlinestatbook.com/stat_sim/sampling_dist/}{Central Limit
Theorem).


## Try at Home

Just to help gain/rehearse intuition about standard errors, try the example
again at home and this time simply make up a new sample size.  That is try
replacing in the calculations above

```{r}
n <- 10
n <- 100
n <- 1000
```

What happens to the width of the confidence interval for each new $n$?


## Take Away

Because math/statistics

* we can approximate confidence intervals for sample means
* we can approximate hypothesis tests (coming soon)...
* t-test shows up more often than normal in CLT
    * their similar in spirit, but t has fatter tails
* for a long time, this was the only way to statistics
* we'll keep referring back to the CLT for the rest of the class
