---
title: "Logistic Regression"
author: "Edward"
format: html
editor: visual
---

```{r}
library(ggplot2)
suppressMessages(library(dplyr))
# normal link: elmhurst -> hospital
hospital <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/hospital.csv")
```

## Linear Regression

```{r}
ggplot(hospital, aes(stay, infection_risk)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
fit <- lm(infection_risk ~ nurses + stay, data = hospital)
summary(fit)
```

```{r}
df <- data.frame(
  yhat = fitted(fit),
  r = rstandard(fit)
)
```

```{r}
ggplot(df, aes(yhat, r)) +
  geom_point()
```

```{r}
ggplot(df, aes(r)) +
  geom_histogram(bins = 9)
```

When the number of nurses is fixed at 50, we expect a one day increase in the number of days stayed in a hospital to increase the infection risk by 0.317.

```{r}
fit %>%
  predict(newdata = data.frame(stay = c(10, 11),
                               nurses = c(50, 50))) %>%
  diff
```

When the number of days in a hospital is fixed at 11, we expect each next nurse to increase the infection risk by 0.002.

```{r}
fit %>%
  predict(newdata = data.frame(stay = c(11, 11),
                               nurses = c(50, 51))) %>%
  diff
```

## Logistic Regression

```{r}
## normal link: hospital -> possum
possum <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/possum.csv")
```

```{r}
ggplot(possum, aes(tailL, pop)) +
  geom_point()
```

```{r}
possum <- possum %>%
  mutate(vic = as.numeric(pop == "vic"))
```

```{r}
fitl <- glm(vic ~ tailL, data = possum, family = binomial())
summary(fitl)
```

```{r}
o <- order(possum$tailL)
df <- data.frame(
  tailL = possum$tailL[o],
  yhat = fitted(fitl, type = "response")[o]
)
```

```{r}
ggplot() +
  geom_point(data = possum, aes(tailL, vic)) +
  geom_line(data = df, aes(tailL, yhat))
```

When tail length increases by 1 cm, from 32 to 33 centimeters, we expect the probability that a possum is from "vic" to go down by 0.04.

```{r}
fitl %>%
  predict(newdata = data.frame(tailL = c(32, 33)), type = "response") %>%
  diff
```

When tail length is increased by 1 cm, from 42 to 43 centimeters, we expect that probability that a possum is from "vic" to go down by 0.01.

```{r}
fitl %>%
  predict(newdata = data.frame(tailL = c(42, 43)), type = "response") %>%
  diff
```













