---
title: "Summaries By Group"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
    code-fold: show
    self-contained: true
tbl-cap-location: bottom
execute:
  echo: true
---

# Basic Idea

## Means by Group

Consider the \texttt{datasets::ChickWeight} dataset.

```{r}
suppressMessages(library(datasets))
# random 6 rows of ChickWeight
ChickWeight[sample(1:nrow(ChickWeight), 6),]
```


## Means by Group

Means of chicks's weight by diet.

```{r}
suppressMessages(library(dplyr))
summarise(group_by(ChickWeight, Diet),
          Mean=mean(weight))
```


## Chicken Weight by Diet, means and more

Mean and standard deviations of chicks's weight by diet.

```{r}
(df <- summarise(group_by(ChickWeight, Diet),
                 Mn = mean(weight),
                 StDev = sd(weight),
                 q1 = quantile(weight, .25),
                 q3 = quantile(weight, .75)))
```


# Workhorses

## Group Summary Statistics, the workhorses

The underlying ideas behind so much of statistics rely on three
ideas/functions and one sentence enhancer.

* **group_by** -- groups a data frame by some variables
* **summarise** -- summarizes many observations into one number
* **mutate** -- mutate variables of a data frame into new variables
* `%>%` -- make code read (almost) like English

## group_by

The function `group_by` is incredibly helpful, but not that exciting.

```{r}
# ?group_by
head(ChickWeight) # start with this
```


## group_by

If we group the dataset ChickWeight by Diet, things change only
slightly.  But what `group_by` returns is now ready to be passed into
`summarise`.

```{r}
group_by(ChickWeight, Diet)
```


## summarise

The function `summarise` collapses multiple observations down into one
number, for instance into a summary statistic.  As we saw before, we
can summarize multiple variables at once.

```{r}
# ?summarise
```

## summarise

What does the following code do?

```{r}
summarise(ChickWeight,
          Mn = mean(weight, na.rm=TRUE))
# hint
mean(ChickWeight$weight, na.rm=TRUE)
```


## summarise

We can also summarize multiple variables at once -- by group or not.

```{r}
summarise(ChickWeight,
          mn = mean(weight),
          sd = sd(weight),
          mdn = median(weight),
          iqr = IQR(weight))
```


## mutate

The function `mutate` allows us to create new variables and add them to
the data frame.  Recall our summarized data named

```{r}
df
```

## mutate

We can create a new variable (column)

```{r}
(newdf <- mutate(df, iqr=q3-q1))
```


## mutate

In case our new variable isn't automatically printed, remember we can make R
print things for us.

```{r}
newdf[,c("Diet", "iqr")]
```


## mutate

`mutate` works on any data frame.  For instance, you might have two variables
that are obviously better off as a ratio.


* miles driven and time (hours) $\Rightarrow$ miles/hr.
* state income and state population $\Rightarrow$ per capita income
* or any state total, $x$, and state population $\Rightarrow$ per capita x
* surface area and volume $\Rightarrow$ surface-area-to-volume ratio, SA:V
* ...


## Group Summary Statistics

Summary statistics and a lesson about missing values in R.

```{r}
library(ape)
data(carnivora)
summarise(
    group_by(carnivora, Family),
        Mn = mean(LS), # litter size
        Total = n(),   # counts
        Sm = sum(LS, na.rm=TRUE))
```

## Group Summary Statistics, take 2

We removed NAs from mean calculation and made code (almost) read like
English

```{r}
carnivora %>%
    group_by(Family) %>%
    summarise(Mn = mean(LS, na.rm = TRUE)) # Litter Size
```

## Group Summary Statistics, take 3

```{r}
carnivora %>%
    group_by(Family) %>%
    summarise(
        Mn = mean(LS, na.rm = TRUE),
        Total = n(),
        Sm = sum(LS, na.rm = TRUE)
    )
```


# My Favorite Example

## mutate

Maybe relative brain size matters to you, i.e. heaviest brain
relative to body weight.

```{r}
heavy <- carnivora %>%
    group_by(Family) %>%
    mutate(brbo = SB/SW) # brain/body
```


## mutate and summarise

Then throw in some `summarise` and find which family has greatest mean
brain to body weight ratio.

```{r}
heavy %>%
    summarise(Mn_brbo = mean(brbo, na.rm=TRUE))
```


## who?

* These guys [Procyonidae](https://en.wikipedia.org/wiki/Procyonidae)
* But also [PNAS article](http://www.pnas.org/content/early/2016/01/20/1505913113)
