---
title: "Exam 01 Practice, solutions"
author: "Edward (if the answer are right, otherwise not Edward)"
format: html
editor: visual
---

1.
```{r}
horse <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/horse_prices.csv")
```

2. observational study, because horse prices were likely not assigned by the researcher

3.
```{r}
suppressMessages(library(dplyr))
library(ggplot2)
```

4.
quantitative/numeric: Price, Age, Height
qualitative/categorical/factor: Sex

5.
explanatory: Age or Height
response: Price

6.
```{r}
ggplot(horse, aes(Price)) + geom_histogram(bins=11)
```
7.
```{r}
mean(horse$Price, na.rm = TRUE)
median(horse$Price, na.rm = TRUE)
```

The histogram looks more or less symmetric, but because of the one
potential outlier near $60,000, the mean is larger than the median.

8.
```{r}
ggplot(horse, aes(Sex, Price)) + geom_boxplot()
```

The median horse price for male horses is higher than the female
median price.  There is one potential outlier for the female horse
prices.  The mean does not show up on box plots.

9.
```{r}
horse %>%
  group_by(Sex) %>%
  summarize(m = mean(Price),
            s = sd(Price),
            n())
```

The variability in horse prices is greater for male horses, than it is
for female horses, as is seen by the larger standard deviation of male
horse prices.  In this dataset, there were 20 female horses and 30
male horses.

10.

  a. The mean and standard deviation for dataset 2. is larger, because
  of the relatively large value 20 pulling up the mean and the
  standard deviation.

```{r}
x <- c(3,5,5,5,8,11,11,11,13)
y <- c(3,5,5,5,8,11,11,11,20)
mean(x)
sd(x)
mean(y)
sd(y)
```


  b. The mean is smaller for dataset 2., because of the relatively
  extreme negative value -40. However, the standard deviation is
  larger for 2., because the -40 is further from the mean in this
  dataset.

```{r}
x <- c(-20,0,0,0,15,25,30,30)
y <- c(-40,0,0,0,15,25,30,30)
mean(x)
sd(x)
mean(y)
sd(y)
```
