---
title: "Paired and Two Sample t-tests"
affiliation: "CSU, Chico MATH 315"
author: "Edward A. Roualdes"
format:
  revealjs:
    incremental: true
    scrollable: true
    self-contained: true
tbl-cap-location: bottom
execute:
  echo: true
---

```{r}
#| include: false
library(ggplot2)

theme_set(
    theme_minimal(base_size = 16)
)
update_geom_defaults("line", list(size = 1.1))
```

# Recap

## Recap: $t$-distribution

So far, we have used the t-distribution for confidence intervals or hypothesis
tests of *one mean* (proportions are means).

```{r}
#| eval: false
t.test(df$x, mu = some_value, conf.level = 0.95)
```

# Paired Data

## Paired Data, definition

**Paired data** are somehow intimately connected.

Two sets of observations are paired if each observation in one set has a special
correspondence or connection with exactly one observation in the other data set.


## Paired Data, by example

Tell me whether or not these data are paired.

* two website's price for the same book
* eye sight ratings by person
* jumping spider vertical (jump height)
* upper verse lower bird beak lengths
* weights of male and female babies


## Paired Data, t-test

If the data are paired, their difference has direct and interpretable meaning
both in English and in statistics; $X_{i,diff} = X_{i,a} - X_{i,b}$ has meaning.
Therefore

$$\bar{X}_{diff} \quad \text{ and } \quad s_{\bar{X}_{diff}}$$

are simply fancy ways to write new random variables.

# An Example

## Paired Data, confidence interval

Are textbooks actually cheaper online?  Compare the price of textbooks at the
University of California, Los Angeles' (UCLA's) bookstore and prices at
Amazon.com.  Seventy-three UCLA courses were randomly sampled in Spring 2010.

```{r}
books <- read.csv("https://raw.githubusercontent.com/roualdes/data/master/books.csv")
## look at data in RStudio
## what plot should we make?
```

# Paired Data, confidence interval

Plot the data!

```{r}
#| eval: false
suppressMessages(library(ggplot2))
ggplot(books, aes(uclaNew, amazNew),
       xlab="UCLA price ($)", ylab="Amazon price ($)") +
    geom_point() +
    geom_abline(intercept=0, slope=1)
```

## Paired Data, confidence interval

Calculate and interpret a 95% confidence interval of the difference in Amazon.com versus UCLA's book prices.

## Paired Data, confidence interval
```{r}
d <- books$amazNew - books$uclaNew # create new variable
# now the same as before
t.test(d, conf.level = 0.95)
```

## Paired Data, confidence interval

```{r}
#| echo: false
#| output: false
xbar <- mean(d)
s <- sd(d)
n <- length(d)
tstar <- qt(0.975, n-1)
xbar - tstar*s/sqrt(n)
xbar + tstar*s/sqrt(n)
```

We are $95$\% confident that the true mean difference in price between Amazon.com and
UCLA's books is between `r round(xbar - tstar*s/sqrt(n), 2)` and
`r round(xbar + tstar*s/sqrt(n), 2)`.

## Paired Data, hypothesis test

Set up, evaluate, and conclude in context a hypothesis test at $\alpha = 0.05$.


## Paired Data, hypothesis test
The natural hypotheses are

$$ H_0: \mu_{diff} = 0 \text{ versus } H_1: \mu_{diff} \ne 0. $$

## Paired Data, hypothesis test

```{r}
d <- books$amazNew - books$uclaNew # create new variable
t.test(d, mu = 0, conf.level = 0.95)
```


## Paired Data, hypothesis test

Becuase p-value$<0.0001 < \alpha = 0.05$, we reject $H_0$.  There is
insufficient evidence to claim that Amazon.com and UCLA's book prices are the
same.

# Two Sample t-test


## Two Sample t-test

Two sample t-tests estimate the difference between two population means from two
independent samples of data.  We estimate $\mu_a - \mu_b$ with the point
estimator $\bar{X}_a - \bar{X}_b$.


## Two Sample t-test, confidence interval

Confidence intervals for two sample t-tests follow the same pattern as before,

$$ (\bar{X}_a - \bar{X}_b) \pm t^*_{df} \cdot s_{\bar{X}_a - \bar{X}_b}. $$

## Two Sample t-test, hypothesis tests

Test statistics for two sample t-tests follow the same pattern as before, and
p-values are exactly the same.



# Two Sample t-test, example


# Two Sample t-test, example

Consider the data set `ape::carnivora`.  Calculate a $98$\% confidence
interval for the difference in mean longevity between the two SuperFamilies
Caniformia and Feliformia.

```{r}
suppressMessages(library(ape))
data(carnivora)
# ?carnivora
# look at importing data
```


## Two Sample t-test, confidence interval

A $98$\% CI, difference in longevity by Caniformia and Feliformia.

```{r}
#| echo: false
model <- t.test(LY~SuperFamily, data=carnivora, conf.level=0.98)
```

```{r}
t.test(LY~SuperFamily, data=carnivora,
       conf.level=0.98)
```


## Two Sample t-test, confidence interval

We are $98$\% confidence that the population difference in mean longevity
between the SuperFamlies Caniformia and Feliformia is between
`r round(model$conf.int[1], 1)` and `r round(model$conf.int[2], 1)`.


## Two Sample t-test, hypothesis test

Set up, evaluate, and conclude in context a hypothesis test at $\alpha = 0.02$.



## Two Sample t-test, hypothesis test

The natural hypotheses are

$$ H_0: \mu_C = \mu_F \text{ versus } H_1: \mu_C \ne \mu_F $$

and $\alpha = 0.02$.


## Two Sample t-test, hypothesis test

Because p-value $=`r round(model$p.value, 2)` > \alpha = 0.02$, we fail to
reject $H_0$.  There is insufficient evidence to claim that the true difference
in mean longevity between Caniformia and Feliformia is different.


# Take Away

## Take Away

Overall things have stayed pretty much the same: confidence intervals,
hypothesis tests, and interpretations.  Now we have new types of data we can
work with.


* Paired data
    * Two variables are intimately connected $\Rightarrow$ their difference has meaning
    * create one variable from the two $\Rightarrow$ one sample t-test

* Two sample data
    * Two variables are independent
    * Point estimate is difference of means
    * Standard error follows from independence
