---
title: "Regression Inference"
format: html
editor: visual
---

```{r}
library(ggplot2)
suppressMessages(library(dplyr))
```


```{r}
elmhurst <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/elmhurst.csv")
```

Want to predict `gift_aid`, measured in \$1,000, using `family_income` also measured in \$1,000.

```{r}
ggplot(elmhurst, aes(family_income, gift_aid)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

There appears to be a moderate, negative, and linear relationship between family income and gift aid.
```{r}
cor(elmhurst$family_income, elmhurst$gift_aid)
```

```{r}
fit <- lm(gift_aid ~ family_income, data = elmhurst)
summary(fit)
```

Fitted regression equation

expected gift aid = 24.32 + -0.04 * family income

or 

$$\widehat{gift\_aid} = 24.32 + -0.04 * family\_income$$
Use ~~multiple~~ Adjusted R^2: 23.29% of the variation in gift aid is explained by this linear regression model of family income.

Intercept: For a family with \$0 in income, we expect their gift aid to be \$`{r} 24.32 * 1000`. 

Slope: For every \$1,000 increase in family income, we expect gift aid to go down by \$`r 0.04 * 1000`.

Prediction: For a family with income of \$200,000, we expect the student to receive \$`r 15.705 * 1000` in gift aid.

```{r}
predict(fit, newdata = data.frame(family_income = 200),         
        interval = "confidence", 
        level = 0.95)
```

Prediction confidence interval: We are 95% confident that for a family with income of \$200,000, we expect the student to receive gift aid between \$`r 13.17 * 1000` and \$`r 18.24* 1000`.


```{r}
confint(fit, level = 0.98)
```

Confidence interval for the intercept: We are 98% confident that for a family with \$0 in income, we expect the student to receive between \$`r 21.21 * 1000` and \$`r 27.43 * 1000`.

Confidence interval for the slope: We are 98% confident that for each next \$1,000 a family earns in income, we expect that amount of gift aid the student receives to go down by between \$`r 0.02 * 1000` and \$`r 0.07 * 1000`.

