---
title: "MATH 456 Practice Exam 01, Spring 26"
format: html
editor: source
---

```{r, echo=FALSE}
library(ggplot2)
suppressMessages(library(dplyr))
df <- read.csv("https://raw.githubusercontent.com/roualdes/data/refs/heads/master/finches.csv")
```

Consider a data set about finches (little birds) from the Galapagos islands.  We'll consider the variables `beakwidth` measured in millimeters (mm), `taillength` also in mm, and `island`.

```{r}
#| echo: false
#| fig-width: 3
#| fig-height: 2
#| fig-align: center
#| warning: false
ggplot(df, aes(island, taillength)) +
   geom_boxplot() +
   theme_minimal()
fita <- lm(taillength ~ island, data = df)
dfa <- as.data.frame(lapply(as.data.frame(anova(fita)), \(x) if(is.numeric(x)) { round(x, 4)} else {x}))
colnames(dfa) <- c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")
rownames(dfa) <- c("island", "Residuals")
dfa$`F value`[2] = "-"
dfa$`Pr(>F)`[2] = "-"
dfa$`Df`[1] = "-"
print(dfa)
summary(fita)
TukeyHSD(aov(fita))
```

1. Write three sentences about three different statistics from the box
   plot above.  Each sentence must be about a different island.

2. Write the null and alternative hypotheses for ANOVA based on the
   box plot above. Specify a level of significance.
   
3. Compare the p-value to the level of significance and make a conclusion.

4. Interpret your conclusion in context of the data.

5. What should the degrees of freedom for `island` be for the ANOVA above?

6. Reproduce the F statistic (named F.value in the output above).

7. Write down the fitted regression equation for the model above.

8. Using Tukey's HSD, write down appropriate null and alternative
    hypotheses for one comparison. Make a conclusion by quoting an appropriate p-value and then interpret your conclusion in context of the data.

```{r}
predict(fita, newdata = data.frame(island = "floreana"))
```

9. Still using the ANOVA model from above, named `fita`, interpret the
   prediction in context of the data.

10. Explain why all finches from the island Floreana have the same
    prediction for this model.

```{r}
#| echo: false
#| fig-width: 3
#| fig-height: 2
#| fig-align: center
#| warning: false
ggplot(df, aes(beakwidth, taillength)) +
   geom_point() +
   theme_minimal()
fitl <- lm(taillength ~ beakwidth, data = df)
summary(fitl)
confint(fitl, level = 0.9)
```

11. Write the null and alternative hypotheses for the hypothesis test
   of the intercept. Specify a level of significance.
  
12. Make the appropriate conclusion from the hypothesis test above. Quote the appropriate p-value.

13. Interpret your conclusion in context of the data.

14. Write down the fitted regression equation for the model above.

15. Interpret adjusted $R^2$ in context of the data.

16. Provide two reasons $R^2$ is worse than adjusted $R^2$?

17. Interpret in context of the data a 90% confidence interval for the
    intercept.

18. Does the intercept make sense in context of the data.  Why or why
    not?

```{r}
predict(fitl, newdata=data.frame(beakwidth=5), interval="confidence")
```

19. Interpret in context of the data a 95% confidence interval for a
    prediction of the mean `taillength` based on the output above.

20. Without doing any calculations, write the regression equation for
    the prediction above.

21. Name two ways to make more narrow the width of a confidence
    interval for a mean `taillength`?

22. Here is pseudo-code for a function that would be passed to
    `optim`, `R`'s built in optimization function, in order to
    calculate linear regression coefficients.  Explain what each line
    of code is doing.

```{r}
ll_lm <- function(theta, data) {
    x <- data$x
    y <- data$y
    yhat <- theta[1] + theta[2] * x
    r <- y - yhat
    return(sum( r ^ 2 ))
}
```
